Palomar 5 Gaps

### 2024-10-11 paper/main.tex
added paolas edits

@@

  The existence of dark matter is supported by substantial observational evidence on scales larger than individual galaxies. This work aims to contribute to the search for dark matter on sub-galactic scales, specifically within the Milky Way. First proposed by [Author Name] in [Year], the concept of dark matter arose to explain the unusually high velocities of stars in dwarf galaxies—velocities that exceed those predicted by the visible mass exerting gravitational forces. Dark matter's influence extends to cosmological scales as well, as demonstrated by the bending of light within cosmic web filaments due to the unseen mass. Despite significant progress, the true nature of dark matter remains elusive, and scientists continue to search for evidence of its existence beyond its gravitational effects on cosmic and extragalactic scales. A key focus today is detecting dark matter on sub-galactic scales, such as through the study of stellar streams.

  Stellar streams, long and thin structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy, present a promising method for probing dark matter distribution. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit. Theoretical predictions of this phenomenon existed long before the first observation in 1995 by \citet{carl_j_grillmair_globular_1995}, who detected stars beyond the tidal radii of globular clusters. Subsequent work by \citet{odenkirchen_detection_2001} further demonstrated that Palomar 5's stars not only extended beyond the tidal radius but also formed distinct tidal tails.

  Over the past two decades, theoretical and numerical studies have shown that stellar streams can constrain the gravitational field of host galaxies, particularly the Milky Way. By applying Poisson's equation, the density distribution and gravitational potential can be separated into different components, including the dark matter halo. This halo, often modeled as spherical, is the largest component and crucial to explaining the constant rotational velocity of galaxies. For example, \citet{varghese2011stellar} demonstrated how stellar streams can infer the mass and scale parameters of haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. Meanwhile, Bonaca et al. employed an information-theoretic approach, identifying the orbits and configurations of stellar streams that yield the most information about the galactic potential.

  The Gaia Data Release 2 (DR2) and DR3 revolutionized the observational study of stellar streams, enabling significant advances in the field. Malhan et al. developed the StreamFinder algorithm, which identified 90 stellar streams within the Milky Way, some of which have been followed up by other instruments and surveys. The best-observed streams, such as GD-1 and Palomar 5, contain between 1,000 and 3,000 stars. Leveraging this dataset, Ibata et al. (2024) constructed the most accurate model of the Milky Way to date based on stellar stream data.

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by Springel et al. (2008), $\Lambda$-Cold Dark Matter ($\Lambda$CDM) predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring, as described by Vingetti et al. (20XX). However, some models predict that dark matter clumps could exist down to the mass of a planet like Jupiter (I hear this at conferences, is it true? where are the papers that suggest this?).

>>>>

\newcommand{\paola}[1]{\textcolor{magenta}{{#1}}}

@@

  \paola{Postulated about a century ago, to explain the large velocity dispersions of galaxies inside clusters \citep[see][]{zwicky33}\footnote{The English translation of the original paper is available at \citet{zwicky_EN}} and the rotation curves of disk galaxies as well \citep{freeman70, rubin70}, different  astronomical measurements conducted over the last century point to the existence of dark matter, from temperature fluctuations in the Cosmic Microwave Background  to weak and strong gravitational lensing \citep[see, among many, ][]{planck20, mandelbaum06, hsueh20}. Dark matter constitutes the dominant ($\sim$~85\%) contributor to matter in the Universe \citep{planck20}, and shapes the building-up of its structures over time \citep{padmanabhan93, bertschinger98, springel05}.}

  Despite significant progress, the true nature of dark matter remains elusive, and scientists continue to search for evidence of its existence beyond its gravitational effects on cosmic and galactic scales \citep[see][ for a review]{arbey21}. A key focus today is detecting dark matter on sub-galactic scales, because at these scales different cosmological scenarios predict different lumpiness of the halos \citep{madau08, zavala19}. \paola{One of the most promising tools to set limit to the halo lumpiness, or granularity, and hence to the number and mass density of the so-called sub-halos, is provided by the study of} stellar streams.

  Stellar streams, several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy, present \paola{indeed} a promising method for probing dark matter distribution. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, \paola{unless in the closest vicinity to the object \citep{montuori07}}. Numerical predictions of this phenomenon exist \paola{since already the '70s of the last century \citep[see, for example][]{keenan75}, well}  before the first detections of \paola{stellar streams around some Galactic globular clusters made by}  \citet{carl_j_grillmair_globular_1995}, \paola{and the nearly contemporary discovery of the Sagittarius stellar stream, the closest (to the Sun) example of a stream emerging from a dwarf satellite currently interacting with the Milky Way \citep{mateo98, newberg02, majewski03, martinez04}.} \paola{Subsequent works \citep[see][]{lehmann97, testa00, leon00, siegel01, lee03} extended Grillmair's findings to other globular cluster streams, until the discovery made by \citet{odenkirchen01,odenkirchen02, odenkirchen03} of long and thin tails outside the Palomar~5 globular cluster, which were shown to be extended for more than $10^\circ$ degrees in the sky.  These early works stimulated a vigorous search for stellar streams in the Galaxy in the following years \citep{grillmair06a, grillmair06b, belokurov06, grillmair09}, which has now entered a new dimension thanks to the Gaia astrometric mission \citep{prousti16}, which allows to look for these structures coupling photometry, astrometry and (when possible) spectroscopic  information. Currently about 20 GCs have been found to possess streams around them \citep[see][ for a summary of these discoveries]{mateu23}, while some streams do not have any clear association with existing GCs, pointing to the early presence of a population of clusters  now completely dissolved in the Galactic field \citep{ibata24, martin22, wan20}.}

  \paola{One of the interests in studying stellar streams, is that they } can constrain the gravitational field of their host galaxies, particularly the Milky Way. For example, \citet{varghese2011stellar} demonstrated how stellar streams can infer the mass and scale parameters of \paola{dark matter} haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \paola{The use of single streams for constraining the Galactic potential has led, however, to some ambiguous and not converging results over time. For example, \citet{law10} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{helmi04, johnston05, law05}; \citet{bovy16} concluded that the dark matter halo of our Galaxy is nearly spherical at the radii of the Palomar~5 and GD~1 streams. } Meanwhile, \citet{bonaca14} employed an information-theoretic approach, identifying the orbits and configurations of stellar streams that yield the most information about the galactic potential.

  \paola{As previously stated, }  the Gaia mission, and its Data Releases 2 and 3 \citep{gaia18, gaia23}, revolutionized the observational study of stellar streams, \paola{unraveling a population of stellar streams much more numerous than that previously known \citep[see, for example][]{malhan18}}, enabling significant advances in the field. Malhan et al. developed the StreamFinder algorithm, which identified 90 stellar streams within the Milky Way, some of which have been followed up by other instruments and surveys. The best-observed streams, such as GD-1 and Palomar 5, contain between 1,000 and 3,000 stars.  Leveraging this dataset, \citet{ibata24} constructed the most accurate model of the Milky Way to date based on a set of about 90 streams.

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by Springel et al. (2008), $\Lambda$-Cold Dark Matter ($\Lambda$CDM) predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring, as described by \citet{vegetti12}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{amorisco21}]{green05, wang20}.
